---
title: "Care about your logarithms"
author: "Euloge Clovis Kenne Pagui and Michele Lambardi di San Miniato"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{numerically sound definitions for distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
require(truncdist)
require(truncLCdist)
```

A a part of defining probability mass/dnesity and cumulative distribution functions, you'll often have to use exponentials and logarithms, and this is a potential source of instability. While $\log(\exp(x))$ is equivalent to $x$, they are not numerically the same.

We rapidly implement the Laplace distribution, with density
$$f(x) = \frac{1}{2\sigma}\exp\left(-\frac{|x-\mu|}{\sigma}\right) \,.$$
We do this by modding the package, see vignette "modding":

```{r}
truncLCdistoptions(
  laplace.continuous=T,
  laplace.m=function(mu=0, sigma=1) mu,
  laplace.d=function(x, mu=0, sigma=1, log=F) {
    y <- -abs((x-mu)/sigma) - log(2*sigma)
    if (log) y else exp(y)
  }
)
```

But what about the cumulative distribution function:
$$F(x) = \frac{\exp\left(-\frac{|x-\mu|}{\sigma}\right)}{2} \text{ if } x < \mu \text{, otherwise } 1 - \frac{\exp\left(-\frac{|x-\mu|}{\sigma}\right)}{2}$$

One may implement it as:

```{r}
plaplace1 <- function(x, mu=0, sigma=1, log.p=F) {
  y <- exp(-abs((x-mu)/sigma))/2
  y <- ifelse(x < mu, y, 1 - y)
  if (log.p) log(y) else y
}
```

But you see, you don't want to first compute the exponential and then the log, the latter being actually used in the package.

The problem is even more relevant in the evaluation of the function
$$\mathrm{log1mexp}(x) = \log(1-\exp(-x)) \,,\quad x > 0 \,.$$
The \texttt{R} package \texttt{Rmpfr} implements it in a computing-aware fashion, which is worth consideration.

```{r}
plaplace2 <- function(x, mu=0, sigma=1, log.p=F) {
  y <- abs((x-mu)/sigma) + log(2) # >= log(2) > 0
  y <- ifelse(x < mu, -y, Rmpfr::log1mexp(y))
  if (log.p) y else exp(y)
}
```

The two versions seem the same:

```{r}
curve(plaplace1(x), from=-3, to=+3)
curve(plaplace2(x), add=T, col="red", lty=2)
```

But they will behave very differently under truncation.

```{r}
a <- 100
truncLCdistoptions(laplace.p=plaplace2)
set.seed(1234)
plot(ecdf(rtruncLC(100000, "laplace", a=a)))
curve(pexp(x-a), add=T, col="red", lty=2)
```

Try with
```{r, eval=F}
truncLCdistoptions(laplace.p=plaplace1)
```
You'll see it hardly works.
