---
title: "Issues with truncdist"
author: "Euloge Clovis Kenne Pagui and Michele Lambardi di San Miniato"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{truncdist: troubles in paradise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
require(truncdist)
require(truncLCdist)
```

The R package truncdist is a Swiss knife for dealing with truncated distributions. The reason for its generality is that it uses inverse tranform sampling, which is a viable option whenever the quantile function is tractable.

The problem with this approach is that it suffers in the tails of a distribution. For instance, a normal distribution truncated between $a=0$ and $b=\infty$ can be sampled this way:

```{r}
n <- 10000
a <- 0
b <- +Inf
set.seed(1234)
summary(rtrunc(n, "norm", a=a, b=b))
```

But what if we are interested in more extreme tails? Say, $a=8$.

```{r}
a <- 8
set.seed(1234)
summary(rtrunc(n, "norm", a=a, b=b))
```

Actually, performing inverse transform sampling manually will exhibit the same issues:

```{r}
set.seed(1234)
summary(qnorm(pnorm(a)+(1-pnorm(a))*runif(n)))
```

This because truncdist just looks up for quantile functions based on names and uses them blindly.

Whenever one needs to simulate extreme events, for weather sciences, or value-at-risk assessments, one has to rely on simulations and the reliability of software is crucial.

This is why an enhanced version of truncdist has been developed that both reimplements its rtrunc routine and provides ways to check whether it is safe to trust the software.
